---
title: "p8105_hw3_fir2103"
author: "Farizah Rob"
date: "2022-10-12"
output: github_document
---

```{r}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6, 
  fig.asp = .6, 
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position="bottom"))

options(
  ggplot2.continuous.colour = "viridis", 
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete <- scale_color_viridis_d
scale_fill_discrete <- scale_fill_viridis_d
```

## Problem 1 

```{r}
library(p8105.datasets)
data("instacart")

instacart <- 
  instacart %>% 
  as_tibble(instacart)
```

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart %>% select(product_id) %>% distinct %>% count` products found in `r instacart %>% select(user_id, order_id) %>% distinct %>% count` orders from `r instacart %>% select(user_id) %>% distinct %>% count` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart %>% 
  count(aisle) %>% 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart %>% 
  count(aisle) %>% 
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle) %>% 
  count(product_name) %>% 
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>% 
  arrange(desc(n)) %>%
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day)) %>%
  spread(key = order_dow, value = mean_hour) %>%
  knitr::kable(digits = 2)
```

## Problem 2

```{r}
accel_data <- read_csv("./data/accel_data.csv") %>% 
  janitor::clean_names() %>% 
  mutate(day_type = as.factor(if_else(
    day == "Saturday"| day == "Sunday", "weekend", "weekday")), 
    day = factor(day, levels = c("Monday", "Tuesday", "Wednesday", 
                              "Thursday", "Friday", "Saturday", 
                              "Sunday")))%>%
  relocate(day_type, .before = activity_1) %>%
    pivot_longer(
    c(activity_1:activity_1440),
    names_to = "activity_min",
    names_prefix = "activity_",
    values_to = "activity_count"
  ) %>%
  mutate(activity_min = as.numeric(activity_min)) %>%
  arrange(day)
```

* The tidied dataset contains `r nrow(accel_data)` rows (observations) and `r ncol(accel_data)` columns (variables). The variables of the dataset are week, day_id, day, weekday (factor variable), activity_min (the minute of the day), activity_count. 


Total activity (24-hour, aggregated all minutes) by day, week
```{r}
accel_data %>% 
  group_by(day, week) %>%
  summarize(total_activity = as.integer(sum(activity_count))) %>%
  pivot_wider(
    id_cols = "day", 
    names_from = "week", 
    names_prefix = "week_", 
    values_from = "total_activity"
  ) %>%
  knitr::kable(digits = 1)
```
* One trend apparent from the table is that  the Saturday from week 3 and week 4 have unusually low activity counts (both are 1440). Monday of week 1 has lower activity count than typical Monday activity counts. 

Single-panel plot for activity counts over the course of the 24-hour period by day 
```{r}
activity_plot <- accel_data %>% 
  ggplot(aes(x = activity_min, y = activity_count, color = day)) +
  geom_point(size = 0.2) + geom_line() +
  scale_x_continuous(breaks=c(0, 180, 360, 540, 720, 900, 1080, 1260, 1440)) + 
  labs(x = "Activity Time", y = "Activity Count", title = "Activity count over 24-hour period")

activity_plot

ggsave("./results/problem_2_plot.pdf")
```

* Between 6 AM and 9 AM (360 and 540 minutes), activity counts for weekdays (Wednesday, Thursday) are higher than weekends.
* Between 9 AM and 12 PM (540 and 720 minutes), activity counts are high - especially on Sundays. 
* Between 12 PM and 4/5 PM, there is a slump in activity counts on almost all days.
* After 6 PM (1080 minutes), there is an increase in activity, especially on Fridays (around 9 PM).

## Problem 3

```{r}
library(p8105.datasets)
data("ny_noaa")

ny_noaa <- ny_noaa %>%
  as_tibble(ny_noaa)

ny_noaa <- ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, c("year", "month", "day"), sep="-", convert = TRUE) %>%
  mutate(month = month.abb[month],
         tmax = as.numeric(tmax),
         tmin = as.numeric(tmin), 
         tmax = tmax/10, 
         tmin = tmin/10,
         prcp = prcp/10)
```

Exploring missing values problem 

```{r}
names(ny_noaa)

ny_noaa %>% 
  count(is.na(tmax))

ny_noaa %>% 
  count(is.na(tmin))

ny_noaa %>% 
  count(is.na(prcp))

ny_noaa %>%
  count(is.na(snow))

ny_noaa %>% 
  count(is.na(snwd))

snow_numbers <- ny_noaa %>%
   group_by(snow) %>%
   summarize(count = n()) %>%
   arrange(desc(count))
```

* The dataset has `r nrow(ny_noaa)` rows (observations) and `r ncol(ny_noaa)` columns (variables). The variables are `id` of the weather station, `month`, `day`, `year`, `prcp` (precipitation in mm), `snow` (in mm), `snwd` (snow depth in mm), `tmax` (maximum temperature in Celsius) and `tmin` (minimum temperature in Celsius). All weather data values are in integer form except `tmax` and `tmin` which are characters. The dataset was tidied by transforming from wide to long format, by changing the activity count variables and `prcp` and `tmin`, `tmax` variables changed.

* There are many missing values of `tmax` and `tmin`. For both `tmax` and `tmin`, about 44% of the values are missing. More than 50% of the `prcp`, `snow`, `snwd` values are non-missing so the main problem arises with the temperature values.

* For snowfall, the most common value is `0`. This is the case since if it snows, it only snows in most areas of New York from the months of November-February/March (therefore, not the majority of the year). 


Two-panel plot showing average max temperature in January and in July in each station across years
```{r}
avg_max_plot <- ny_noaa %>% filter(month %in% c("Jan","Jul")) %>% 
  group_by(month, year, id) %>% 
  summarize(
    tmax_mean = mean(tmax, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = year, y = tmax_mean, group = id)) +
  geom_line(alpha = 0.5) +
  facet_grid(. ~ month) +
  labs(y = "Average Maximum Temperature", title = "Average Maximum Temperature in January vs July")

avg_max_plot

ggsave("./results/problem_3_plot_1.pdf")
```
* Average temperatures in January are obviously much lower than July - most average temperatures across stations are within 10 degrees Celsius of each other. The trends are pretty constant more or less. 

* There is an outlier in July 1986/1987 where the average maximum temperature at one station (15 degrees Celsius) was lower than the typical trend. There is a slight outlier around 1982 where the average maximum temperature was around -12 (lower than typical).


Two-panel plot
```{r}
tmax_tmin_plot <- ny_noaa %>%
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex(bins=40) + labs(x = "Minimum Temperature", 
                           y = "Maximum Temperature", 
                           subtitle = "Hexagonal Heatmap of Min Temp vs Max Temp")

snow_plot <- ny_noaa %>% 
  filter(snow > 0 & snow < 100) %>%
  mutate(year = as.factor(year)) %>%
  ggplot(aes(x = snow, color = year)) + 
  geom_density() + theme(legend.key.size = unit(2, 'mm')) + 
  labs(subtitle = "Density Plot of Snowfall (mm) over 1981-2010")

temp_snow_plot <- tmax_tmin_plot + snow_plot

temp_snow_plot

ggsave("./results/problem_3_plot_2.pdf", temp_snow_plot, width = 10, height = 7)
```


